Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
finished structure rewamp. Now printing and making extra pass on the whole thing

\begin{figure*}[ht!]
		\centering
		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
		\caption{Runtime Memory and Simulated Counter-Part}
		\label{fig:SimulatedMemory}
\end{figure*}

@@

\todo{Shorten and abstract next paragraph. }

The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.  As code is JITted and specific variables and runtime routines are referenced the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will attempt to raise an exception.  Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes a Smalltalk method, handleSimulationTrap:, that invokes the relevant closure.  Once the closure has completed, the processor simulator is advanced to the nextpc and simulation resumes.  In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.

@@

\todo{Rework. maybe check interaction with section JIT debug.}

@@

\todo{Merge 1 and 2 + some shortening}

%1

Each processor simulator is expected to be able to disassemble its machine code.  The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.  For example, here is a snippet of decorated disassembly for x64:

@@

\begin{lstlisting}
00002063: movq 0x800(%rbx), %rax = 'stackLimit' : 48 8B 83 00 08 00 00 
0000206a: cmpq %rax, %rsp : 48 39 C4 
0000206d: jb .-0x47 (0x2028=indexOf:startingAt:ifAbsent:@28) : 72 B9 
HasBytecodePC bc 40/41:
0000206f: movq start@20(%rbp), %rsi : 48 8B 75 14 
00002073: movq anElement@28(%rbp), %rdi : 48 8B 7D 1C 
00002077: movq self@-12(%rbp), %rdx : 48 8B 55 F4 
0000207b: movq $0x0, %rcx : 48 C7 C1 00 00 00 00 
00002082: call .-0x1B2F (0x558=ceSend2Args) : E8 D1 E4 FF FF 
IsSendCall indexOf:startingAt: bc 44/45:
00002087: movq %rdx, index@-20(%rbp) : 48 89 55 EC 
\end{lstlisting}

\todo{RM end of next paragraph or ref}

This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.  \emph{\%rbx} is used as a base register pointing at the variables in interpreter.  Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.  HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc. These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.  This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.

%2

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.  Further, a particular method that exercises some new part of the JIT may be problematic to have executed suring simulation and hence not easily JITted during simulation.

To work around these problems, we implemented a tool called \emph{In-image compilation}. In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e. \emph{not} in the simulated heap) to generate the corresponding machine code. In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text window. Indeed the above snippet is from an in-image compilation.  To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulaiton byte array.  This includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

\begin{figure}[ht!]
		\centering
		\includegraphics[width=\linewidth]{figures/InImageCompiler}
		\caption{In-image compilation}
		\label{fig:InImageCompiler}
\end{figure}

\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes. We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and  
a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.

Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecoded method from the assembly code.  Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.

@@

\todo{add intro about this is anecdote and examples of how we used it ; fix section }

@@

\todo{Remove Assertion references}
\todo{Make it more like a story like the following 2 subsec}
Recently, to evaluate new old space garbage collection algorithms we designed against standard algorithms, we implemented a Mark-Sweep in addition to the existing Mark-Compact collector. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was kind of working out of the box. To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

@@

\paragraph{Simulation.} Once we had partially written the new Sweep algorithm, we started the simulator. Since the algorithm was partially written, we had to write the missing pieces inside the debugger, installing the new code at runtime, as one can do it in any Smalltalk application. In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory (\emph{i.e.,} the heap), performs the GC there, and if no assertion fails, it then performs the GC on the original version. This means that if the algorithm was not working, and likely an assertion would fail or an error such as incorrect memory access would be raised, the erratic behavior would have happened in a copy of the heap. The simulator can then reproduce the exact same erratic behavior as many times as the VM developer wants duplicating again and again the same original heap. This is very convenient to debug specific GC bugs. Indeed, often in GC bugs, when the bug happens, the memory is already corrupted and it's quite difficult to track back where the bug comes from without rewinding the memory state, which is usually very difficult or impossible.

@@

Executing the faulty method led to an assertion failure. However, that assertion failed in a GC store check, telling us that the object to store into looked suspicious (address outside of the heap). It happens that this object was read from a field on stack, and that this field held an incorrect address. We could not tell anymore at this point in the execution what instruction among the thousands previous ones wrote on stack the invalid address. So we discarded the cloned simulation, and cloned a fresh simulator again, just before the execution of the faulty method in machine code to reproduce again the crash. This time, we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to the incorrect value found before. In this case, the conditional breakpoint is checked in between each machine instruction, and execution stopped right after the machine instruction which wrote the incorrect value on stack. From the machine instruction address, we could figure out which bytecode pattern generated the incorrect machine code (it was the new bytecode template for inlined allocations). From there, we built a simpler method crashing the runtime and fixed the template using in-image compilation (See next subsection).

@@

\todo{Make following 2 subsections more like a story like the previous ones}

A side-effect of VM simulation, and specifically to be able to interrupt the simulation and introspect the simulated memory and simulation specific objects, is to be able to analyse the runtime with scripts written on-the-fly. Indeed, the simulator can be stopped at any given point and arbitrary Smalltalk code can be written and evaluated similarly to the \emph{eval} Javascript construct to analyse the simulated memory, including the machine code zone, the heap or any Smalltalk object representing the VM state.

@@

The results of the analysis are sometimes used to direct performance design decisions on the VM. In this section we describe how the analysis impacted a design called "Early polymorphic inline cache promotion".

@@

@@

\todo{Merge 1 and 2}

%1
\paragraph{Interpreter-only simulation.} 
 Since executing machine code on the processor simulators is not fast, the simulated interpreter may be faster \footnote{but JIT code is so much more optimized than bytecode interpretation that it may indeed be faster, depending on the code being executed}, and startup time and memory footprint are smaller.

 %2

@@

\paragraph{Performance.}
The first limitation is due to the simulation performance. The interpreter-only is simulator is around 200 times slower to execute code than the normal VM. With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM. We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.

This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator. Bugs in the jitted code are worse. Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs. In general, once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes. The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds. If the bug is unrelated to the JIT, the interpreter-only simulator can also be used and it is a little bit quicker to execute code.

@@

Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. This section focuses on the two closest related works.

\todo{We need to mention something about graal, maybe high level lang for VM and debugging in general then refs}

\paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18. They allow to inspect the running state of the Maxine VM while it runs for debugging purposes. One of the main difference with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode. We believe having two different modes allows us to easily generate a production VM while still having nice debugging features. Having a full metacircular VM would be interesting, but it is unclear whether it is convenient to build a production VM in that way. So far, most production VMs (Java, Javascript, etc.)are still compiling through the C/C++ compiler and are not metacircular

>>>>

\begin{figure*}[ht!]
		\centering
		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
		\caption{Runtime Memory and Simulated Counter-Part\vspace{-0.2cm}}
		\label{fig:SimulatedMemory}
\end{figure*}

@@

%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects. As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.  %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  
The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
the relevant closure. Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes. In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.

@@

@@

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes, in the simulator, around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid. In addition, when developing a specific part of the JIT, it is convenient to compile a well-chosen simple method exercising that specific new part. This is difficult to do with the full simulator since that new part of the JIT might be called by the start-up logic, leading to the erratic behavior before expected.

To work around these problems, we implemented a tool called \emph{In-image compilation}. In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e. \emph{not} in the simulated heap) to generate the corresponding machine code. In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code, decorates the disassembly and then dumps the output in a text window.

To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulation byte array.  This includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the in-image compilation process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

\begin{figure}[th!]
		\centering
		\includegraphics[width=\linewidth]{figures/InImageCompiler}
		\caption{In-image compilation\vspace{-0.2cm}}
		\label{fig:InImageCompiler}
\end{figure}

@@

\begin{figure}[th!]
		\centering
\begin{lstlisting}
00002063: movq 0x800(%rbx), %rax = 'stackLimit' : 48 8B 83 00 08 00 00 
0000206a: cmpq %rax, %rsp : 48 39 C4 
0000206d: jb .-0x47 (0x2028=indexOf:startingAt:ifAbsent:@28) : 72 B9 
HasBytecodePC bc 40/41:
0000206f: movq start@20(%rbp), %rsi : 48 8B 75 14 
00002073: movq anElement@28(%rbp), %rdi : 48 8B 7D 1C 
00002077: movq self@-12(%rbp), %rdx : 48 8B 55 F4 
0000207b: movq $0x0, %rcx : 48 C7 C1 00 00 00 00 
00002082: call .-0x1B2F (0x558=ceSend2Args) : E8 D1 E4 FF FF 
IsSendCall indexOf:startingAt: bc 44/45:
00002087: movq %rdx, index@-20(%rbp) : 48 89 55 EC 
\end{lstlisting}
		\caption{Decorated disassembled code\vspace{-0.2cm}}
		\label{fig:ASMDecorated}
\end{figure}

\paragraph{Decorated Disassembly.}
Each processor simulator is expected to be able to disassemble its machine code.  The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.  For example, Figure \ref{fig:ASMDecorated} is a snippet of decorated disassembly for x64.

This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.  \emph{\%rbx} is used as a base register pointing at the variables in interpreter.  Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.  HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc. These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.  This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs for debugging purposes. %to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.

\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes. We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates. Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method. Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.

@@

In this section we describe our personal experience using the simulation infrastructure to develop the VM. The different subsection consists in different tasks we had to do, emphasising how we used the simulator and how it helped us perform the each given task.

@@

Recently, to evaluate a new old space garbage collection algorithms, we decided to implement standard algorithms to get reference benchmark results. Since we had already a Mark-Compact collector, we mainly introduced a Mark-Sweep non moving GC as an alternative. The whole implementation was done in the simulator, and only when it was working there, was it compiled to C. Using this process, the compiled C code was working out of the box. %To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

We started by sketching the new Sweep algorithm in the code base. Once it was partially written, we started the simulator. Since the algorithm was not complete, we added the missing pieces inside the debugger, installing the new code at runtime and checking against the current values of the function parameters if the function behaved as expected, as one can do it in any Smalltalk application.

In addition, the simulator has an interesting property: each time it performs a GC, either a scavenge or an old space collection, it first copies the simulated memory, performs the GC there, and if no exception was raised\footnote{The VM code base includes many assertions to ensure specific state is correct at given points in the execution, such assertions are optionally compiled in the production VM, they represent the majority of exceptions raised during development or bug resolution.}, it then performs the GC on the original version. This means that each time the algorithm did not work correctly, the erratic behavior happened in a copy of the heap. Since we can clone again the original version as many times as we want, we were able, each time the memory got corrupted beyond repair, to restart the exact same GC execution on the exact same heap with each objects at the exact same addresses. Indeed, in most cases when the bug happens, the memory was already very corrupted and it was very difficult to track back where the bug came from without rewinding the memory state. We reproduced the exact same GC many times in different cases to fully understand specific bugs.

@@

@@

Executing the faulty method led to an assertion failure. However, that assertion failed in a GC store check, telling us that the object to store into looked suspicious (address outside of the heap). It happens that this object was read from a field on stack, and that this field held an incorrect address. We could not tell anymore at this point in the execution what instruction among the thousands previous ones wrote on stack the invalid address. So we discarded the cloned simulation, and cloned a fresh simulator again, just before the execution of the faulty method in machine code to reproduce again the crash. This time, we enabled single stepping (\emph{i.e.,} the processor simulator simulates one instruction at a time) and we added a breakpoint stopping execution when the specific field on stack would be written to the incorrect value found before. In this case, the conditional breakpoint is checked in between each machine instruction, and execution stopped right after the machine instruction which wrote the incorrect value on stack. From the machine instruction address, we could figure out which bytecode pattern generated the incorrect machine code (it was the new bytecode template for inlined allocations). From there, we built a simpler method crashing the runtime and fixed the template using in-image compilation.

@@

The simulator can be stopped at any given point and arbitrary Smalltalk code can be written and evaluated similarly to the \emph{eval} Javascript construct. We used this technique to analyse the simulated memory, including the machine code zone, the heap or any Smalltalk object representing the VM state.

@@

The results of specific analysis are sometimes used to direct performance design decisions on the VM. In this section we describe how the analysis on polymorphic inline caches impacted a design called "Early polymorphic inline cache promotion".

@@

In this section we first discuss the VM simulator limits and how we work around them, then the interaction between the simulator and bootstrapping processes. Lastly, we mention some relevant related works.

@@

@@

\paragraph{Performance.}
The first limitation is due to the simulation performance. The interpreter-only is simulator is between a hundred and a thousand times slower to execute code than the normal VM. With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation performance usually drops by a factor of two, though in specific cases it is actually faster. We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.

In practice, if a GC bug happens in an application 15 minutes after start-up, it takes around 50 hours to reproduce in the simulator. Fortunately, we work around this problem by using snapshots. Once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes. The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds. In addition, since the interpreter-only simulator is usually quicker at executing code, if the bug is unrelated to the JIT (typically, a GC bug), the interpreter-only simulator can be used.

@@

Many VM developers implemented different tools to help them working more efficiently on their VM, but they rarely publish about it. For example, the teams working on the Truffle/Graal projects \cite{OneVMToRule} have very convenient development tools for their VMs such as a tool to visualize the intermediate representation of their JIT Graal at different point in the compilation process. But, despite an impressive number of publications on the projects, including publications on development tools for the guest languages \cite{TrufflePolyglotDebugger} or on specific parts of the VM \cite{GraalIR2}, no publication seems to be directly related to the VM development tools.

This section discusses some related work: the Maxine inspector and the RPython toolchain infrastructure.

\paragraph{Maxine Inspectors.} The Maxine inspectors \cite{MaxineInspector} were demonstrated at OOPSLA'18. They allow to inspect the running state of the Maxine VM while it runs for debugging purposes. One of the main difference with our design is that the Maxine VM is metacircular, hence it does not have a simulation and a production mode as we do but a single production debuggable mode. We believe having two different modes allows us to easily generate a production VM while still having nice debugging features. Having a full metacircular VM would be interesting, but it is unclear whether it is convenient to build a production VM in that way. So far, most production VMs (Java, Javascript, etc.) are still compiling through the C/C++ compiler and are not metacircular.
